---
title: "GeoBench: Spatial Data Benchmarks (NZ)"
format: gfm
---

Benchmarks for spatial data operations using the `spData::nz` dataset (New Zealand regions).

## Source Code

-   [sf (R)](scripts/bench_r.R)
-   [geopandas (Python)](scripts/bench_py.py)
-   [sedonadb (Python)](scripts/bench_sedona_py.py)
-   [sedonadb (Python + Polars)](scripts/bench_sedona_polars.py)
-   [sedonadb (R)](scripts/bench_sedona_r.R)
-   [duckdb (R)](scripts/bench_duckdb_r.R)
-   [duckdb (Python)](scripts/bench_duckdb_py.py)

## Experimental Setup

We use the New Zealand regions (16 polygons) and generate 100,000 random points within the bounding box of New Zealand.
The primary operation benchmarked is a **Spatial Left Join**: joining the region `Name` to each point.

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(spData)
library(ggplot2)

data(nz)
set.seed(42)
# Sample 100 points for visualization
bbox_poly <- st_as_sfc(st_bbox(nz))
points_sample <- st_sf(geometry = st_sample(bbox_poly, size = 100))

ggplot() +
  geom_sf(data = nz, fill = "lightblue", color = "white") +
  geom_sf(data = points_sample, color = "red", alpha = 0.5, size = 1) +
  scale_x_continuous(labels = function(x) paste0(x / 1000, " km")) +
  scale_y_continuous(labels = function(y) paste0(y / 1000, " km")) +
  coord_sf(datum = st_crs(nz)) +
  theme_minimal() +
  ggtitle("Experimental Setup (n=100)")
```

## Benchmark Results

Run on 100,000 points and 16 polygons (Regions).

- **sf / geopandas / sedonadb**: Standard benchmarks via GPKG.
- **duckdb-gpkg**: DuckDB reading via standard GDAL/OGR (simulates traditional workflow).
- **duckdb-parquet**: DuckDB reading optimized GeoParquet (simulates modern native workflow).

[Download full results (CSV)](results.csv)

```{r plot-results, echo=FALSE, message=FALSE, warning=FALSE}
if (file.exists("results.csv")) {
  library(ggplot2)
  results <- read.csv("results.csv")
  results$system <- factor(results$system,
                           levels = c("sf",
                                      "geopandas",
                                      "sedonadb-sf",
                                      "sedonadb-pandas",
                                      "sedonadb-polars",
                                      "duckdb-gpkg",
                                      "duckdb-parquet"),
                           ordered = TRUE)

  ggplot(results, aes(x = system, y = ops_per_sec, fill = system)) +
    geom_col() +
    facet_wrap(~ operation, nrow = 1, scales = "free_y") +
    scale_fill_brewer(palette = "Set2") +
    theme_minimal() +
    labs(title = "Benchmark Results",
         y = "Operations per second", x = "System") +
    theme(legend.position = "bottom",
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
}
```

## Running the Benchmarks

This repository includes a Docker submodule designed to handle the dependencies for R (sf, duckdb, sedonadb) and Python (geopandas, sedona, polars).

**1. Build the Docker Image:**

```bash
cd docker_submodule
docker build -t geobench .
```

**2. Run the Benchmarks:**

```bash
cd ..
docker run --rm -ti --net=host -v ${PWD}:/home/rstudio/project geobench /bin/bash -c "cd /home/rstudio/project && ./scripts/run_all.sh"
```

**3. Render/Update README.md:**

```bash
docker run --rm -ti --net=host -v ${PWD}:/home/rstudio/project geobench /bin/bash -c "cd /home/rstudio/project && quarto render README.qmd"
```

### Performance Hypothesis

The results show `sedonadb-sf` (R) outperforming Python variants in spatial operations, particularly joins. We hypothesize this is due to efficient, low-overhead data transfer between R and the underlying Sedona/DataFusion engine (likely via `nanoarrow` or C-level pointers), whereas the standard Python path incurs significant cost deserializing WKB into Python Shapely objects. The `sedonadb-polars` workaround bridges this gap significantly by keeping data in Arrow/WKB format, but R's integration currently appears more seamless for these workloads.

The inclusion of DuckDB allows us to compare "native" execution against standard WKB-based approaches:

1.  **Serialization Overhead**: `geopandas` and `sf` (reading GPKG) rely on OGR/GDAL, which requires parsing WKB (Well-Known Binary) into in-memory geometry objects. This deserialization step is often the bottleneck.
2.  **Native Geometry**: `duckdb-parquet` utilizes the spatial extension's ability to read geometry data directly from Parquet without expensive WKB conversion. We hypothesize this will offer significant speedups over the `duckdb-gpkg` approach, effectively isolating the I/O and serialization cost from the actual spatial join compute time.

### Next Steps

1.  **Memory Profiling**: Analyze memory usage to verify object overhead in Python vs R.
2.  **Scale Testing**: Run benchmarks on larger datasets (e.g., >1 million points) to assess scalability.
3.  **Python Optimization**: Develop native Polars/Arrow integration in `sedona-db` Python bindings to eliminate serialization bottlenecks.
4.  **Engine Comparison**: Deep dive into the performance differences between Sedona's Rust-based engine and DuckDB's vectorized execution engine for spatial joins.

**Note**: These benchmarks are intended for illustrative purposes and may not be fully representative of complex, real-world use cases or performance at larger scales. For more comprehensive benchmarks comparing Sedona, DuckDB, and GeoPandas, we recommend checking out the official [Sedona Spatial Benchmarks](https://sedona.apache.org/spatialbench/single-node-benchmarks/).
