---
title: "GeoBench: Spatial Data Benchmarks (NZ)"
format: gfm
---

Benchmarks for spatial data operations using the `spData::nz` dataset (New Zealand regions).

## Source Code

-   [sf (R)](scripts/bench_r.R)
-   [geopandas (Python)](scripts/bench_py.py)
-   [sedonadb (Python)](scripts/bench_sedona_py.py)

## Experimental Setup

We use the New Zealand regions (16 polygons) and generate 100,000 random points within the bounding box of New Zealand.
The primary operation benchmarked is a **Spatial Left Join**: joining the region `Name` to each point.

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(spData)
library(ggplot2)

data(nz)
set.seed(42)
# Sample 100 points for visualization
bbox_poly <- st_as_sfc(st_bbox(nz))
points_sample <- st_sf(geometry = st_sample(bbox_poly, size = 100))

ggplot() +
  geom_sf(data = nz, fill = "lightblue", color = "white") +
  geom_sf(data = points_sample, color = "red", alpha = 0.5, size = 1) +
  theme_minimal() +
  ggtitle("Experimental Setup (n=100)")
```

## Benchmark Results

Run on 100,000 points and 16 polygons (Regions).

```{r results-table, echo=FALSE, message=FALSE}
if (file.exists("results.csv")) {
  results <- read.csv("results.csv")
  knitr::kable(results)
} else {
  print("No results found yet.")
}
```

```{r plot-results, echo=FALSE, message=FALSE, warning=FALSE}
if (file.exists("results.csv")) {
  library(ggplot2)
  results <- read.csv("results.csv")
  
  ggplot(results, aes(x = system, y = time_sec, fill = language)) +
    geom_col() +
    facet_wrap(~operation, scales = "free_y") +
    theme_minimal() +
    labs(title = "Benchmark Results by Operation",
         y = "Time (seconds)", x = "System") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```